Are environmental and geographic variables effective surrogates for genetic variation in conservation planning?

### 2016-01-30 article/article.Rmd
+ cost data + create multi-species priortisation with cost

## Introduction
<!--  main text -->
Biodiversity is under threat. Over the last few decades, many species have gone extinct [@r270; @r272]. Today, thousands of species are listed as endangered or near threatened in the International Union for Conservation of Nature (IUCN) red list (\url{www.iucnredlist.org}). In the future, species are expected to face even more threats [eg. climate change; @r236; @r243]. In response to this crisis, the overarching aim of conservation is to maximize the long-term persistence of biodiversity [@r255; @r67].

The act of setting aside areas for preservation is the hallmark of modern conservation [@r440]. Protected areas buffer species from anthropogenic impacts and provide places for biodiversity processes to persist (eg. nutrient cycling and water purification). Additionally, protected areas set the stage for direct management actions (eg. habitat rehabilitation, translocation, and invasive species management). Since the resources available for conservation are limited, the plans for protected areas (prioritizations) are often generated using spatially explicit biological data (eg. species distribution data) and decision support tools [eg. $\texttt{MARXAN}$; @r22; $\texttt{Zonation}$; @r429] that identify cost-effective solutions.

@@

Recently, conservation scientists have begun investigating the use of surrogates for genetic data to generate more effective prioritizations without the overheads associated with genetic data. Since adaptive genetic variation is driven by selection pressures, environmental variables have been explored as potential surrogates for genetic variation [eg. @r247; @r2]. On the other hand, neutral variation arises from a reduction in gene flow, and so variables that predict reduction in gene flow between areas have been explored. For instance, the isolation by distance theory [IBD; @r259] predicts that populations further apart will have lower gene flow, and in turn have greater neutral variation among them. The use of distance-based surrogates has been found to improve prioritizations in an island network [@r200]. However the effectiveness of distance-based surrogates remains untested in systems where connectivity is complicated by additional factors (eg. elevation in terrestrial systems). Before conservation planners can reliably use surrogates for genetic data in conservation planning, the effectiveness of these surrogates needs to be verified. Otherwise, conservation planners may waste precious resources on securing places that do little to benefit biodiversity.

@@

## Methods
### Study system
To address the aims of the study, we utilized species distribution and genomic data collected by the IntraBioDiv project [@r451; see @r452 for methodological details]. This dataset was well suited for this study, because it provides spatially explicit genomic data for a multitude of species across a broad geographic area and over large environmental gradients. As part of this project, data for 27 alpine plant species were collected across the European Alps according to a grid (22.3 km $\times$ 25 km; Figure S1). Briefly, in every second cell, plant samples were collected from three individuals for each species present inside the cell. Cells that did not contain any plant samples were omitted from our analysis. Samples were genotyped using amplified fragment length polymorphisms [AFLP; @r453]. Matrices denoting the presence/absence of polymorphisms at loci were constructed independently for each species. Thus for each second grid cell, for each of the species in the grid cell, the dataset contains information describing the genomic variation of individuals inside the grid cell. We used these grid cells as planning units to develop prioritizations to investigate the effectiveness of surrogates for the (AFLP) genetic variation. Most of our analyses described hereafter were performed using the \texttt{R} software package [version `r paste0(version$major, '.', version$minor)`; @r192].

### Surrogate data
We explored the effectiveness of climatic and geographic surrogates. To describe variation in the location of each grid cell, we projected the grid into an equi-distant coordinate system ($\texttt{ESRI:102031}$) and used the centroid of each cell. To describe the climatic variation among each cell, we obtained 21 bioclimatic variables [approx. 1 km resolution; @r33]. These layers and the grid cells were projected into an equal-area coordinate system ($\texttt{ESRI:102014}$). The bioclimatic variables were clipped to the extent of the grid and subject to a principle components analysis (PCA; Table S1). The first two principle components (PCs) cumulatively explained `r round(pca.DF[[4]][2],1)` % of the variation and were used to generate two new layers. We calculated the average value for each PC layer in each grid cell, and used these values to characterize the main themes of climatic variation among the grid cells (Figure S2).

### Outlier locus detection
To investigate the effectiveness of the surrogates for adaptive and neutral genetic variation, we first identified which of the sampled loci for each species were under selection. Specifically, we used outlier locus detection methods [evaluated in @r454] to avoid circularity issues. The basic premise underpinning these analyses is that neutral loci are expected to exhibit a specific level of variation among populations ($F_{st}$), and loci that deviate from this expectation should be under selection. In an earlier study, Bothwell _et al._ [-@r455] applied such outlier detection methods to one of the species in this dataset (_Getiana nivalis_). Here, we adopted a similar methodology and applied it to each of the 27 species in the dataset.

@@

### Prioritizations
We used the unreliable representative and adequate prioritization (URAP) formulation of the reserve selection problem [@r465]. This formulation can identify the minimum number of planning units required to preserve a proportion of the species' range as well as a proportion of the species' variation using an attribute space [_sensu_ @r465]. We treated the genomic and surrogate variables as distinct attribute spaces. We set the demand points for each species and its attribute space using the values associated with the planning units. Although Hanson _et al._ [-@r465] recommend against using the same coordinates for demand points and planning units for real-word conservation planning exercises, we did this to provide a more robust test and reduce uncertainty in the analysis. For every prioritization we generated, we calculated the proportion of the adaptive (if detected) and neutral genetic variation that every prioritization secured to assess its performance. We solved all reserve selection problems to within `r (gurobi.params.LST[[MODE]]$MIPGap*100)` % of optimality using the $\texttt{rapr}$ R package [@r465] and $\texttt{Gurobi}$ [@r463].

To understand how prioritizations generated using surrogate-based targets might perform under typical conservation planning scenarios, we generated a set of single-species and multi-species prioritizations. All of these prioritizations used `r (rapr.params.LST[[MODE]]$amount.target*100)` % amount-based targets to ensure an adequate proportion of habitat was secured for each species [@r197; @r255; @r471]. We generated a set of single-species prioritizations for each species: `r rapr.params.LST[[MODE]]$single.species.amount.replicates` prioritizations using only the amount-based targets, a single prioritization using the amount-based target and `r (rapr.params.LST[[MODE]]$surrogate.target*100)` % surrogate-based targets, and a single prioritization using the amount-based target and `r (rapr.params.LST[[MODE]]$genetic.target*100)` % genetic-based targets. We also generated three multi-species prioritizations using the three unique combinations of targets used to generate the single-species prioritizations. Note that we obtained multiple solutions for the amount-based single-species prioritizations because, unlike the other prioritizations, these targets yielded many optimal solutions with varying levels effectiveness.

@@

## Statistical analysis
We tested for differences in the effectiveness of prioritizations representing typical scenarios using generalized linear mixed-effects models [GLMMs; using the $\texttt{lme4}$ R package; @r469]. We fit two full models to the proportion of genomic variation secured in the single- and multi-species prioritizations using a logit link. Both models contained a predictor variable indicating if the prioritization was generated using amount, surrogate, or genetic targets, a second predictor variable indicating if the genetic variation was adaptive or neutral, and the interaction between these variables. Additionally, both models contained an intercept-only random effect to ensure that comparisons were made within species. Each full model was subject to a step-wise term deletion routine using likelihood ratio tests (LRT) to assess term significance (terms were retained when $P < 0.05$). Finally, we conducted a post-hoc analysis using Tukey constrasts with a Bonferroni correction [using the $\texttt{multcomp}$ R package; @r470].

We investigated the relationship between surrogate-based targets and the genetic variation secured in a prioritization using nonlinear mixed-effects models (NLMMs; using the $\texttt{lme4}$ R package). Two models were fit to the adaptive and neutral genetic variation secured using a negative exponential formulation to estimate the rate of the accumulation of genetic variation ($y = 0 + (0 - 1) \times exp(-r \times x)$; $r$ is the accumulate rate). Both models accomodated random effects for the accumulation rate ($r$) parameter.

@@

<!-- Multi-species prioritizations -->
Surrogate-targets were not found to improve the performance of multi-species prioritizations (Figure 2; Figure 3). No significant differences in the effectiveness of prioritizations were found depending on whether they were generated using different targets (${\chi^{2}}_{`r multi.spp.AOV3[['Df']][1]`,`r multi.spp.AOV3[['Df']][2]`} = `r round(multi.spp.AOV3[['Chisq']][2],2)`$; $P `r pretty.pval(multi.spp.AOV3[['Pr(>Chisq)']][2])`$), if they were assessed using adaptive or neutral genetic variation (${\chi^{2}}_{`r multi.spp.AOV2[['Df']][1]`,`r multi.spp.AOV2[['Df']][2]`} = `r round(multi.spp.AOV2[['Chisq']][2],2)`$; $P `r pretty.pval(multi.spp.AOV2[['Pr(>Chisq)']][2])`$), nor the interaction between these variables (${\chi^{2}}_{`r multi.spp.AOV1[['Df']][1]`,`r multi.spp.AOV1[['Df']][2]`} = `r round(multi.spp.AOV1[['Chisq']][2],2)`$; $P  `r pretty.pval(multi.spp.AOV1[['Pr(>Chisq)']][2])`$).

<!-- Correlation analysis -->
The surrogate variables showed a greater utility for neutral genetic variation than adaptive variation (Figure 4). The accumulation rate ($r$) was `r round(summary(env.correlation.NLMM)$coefficients[1],2)` $\pm$ `r round(summary(env.correlation.NLMM)$coefficients[2],2)` S.E. for the proportion of adaptive variation captured in a prioritization using environmental surrogates. On the other hand, the rate ($r$) was `r round(summary(geo.correlation.NLMM)$coefficients[1],2)` $\pm$ `r round(summary(geo.correlation.NLMM)$coefficients[2],2)` S.E. for neutral variation using geographic surrogates. Thus minor increases in the target for either surrogate results in major increases in the benefit of the prioritization. However, the coefficient was much larger for neutral genetic variation so minor increases in the target can acheive substantial gains.

## Discussion
### Introduction
* aim
	+ we aimed to investigate the benefits of using surrogates for genetic variation in conservation planning.
* innovation
	+ first to test the effectiveness of environmental and geographic surrogates in a conservation planning context
* key findings
	+ surrogate variables are beneficial in single-species contexts.
	+ but not in multi-species contexts when a comprehensive set of species is used.
* consequences
	+ highlights the benefit of using a comprehensive set of species
	+ supports literature on using comprehensive set of species

### Single + multi-species analysis

### Correlation analysis

### Limitations/Future directions
* spatial/surrogate data
	+ isolation by resistance: least-cost paths and circuit theory distances are becoming more population for modeling connectivity, and may offer better surrogates than linear geographic distances. However, the scaling for these variables requires an understanding of the often non-linear relationship, which is not available to conservation planners
	+ cost data: here we used no cost data to provide a rigorous test for the benefits of surrogates in conservation planning. however in real conservation planning situations, prioritizations are largely driven by cost. future studies should look at including cost data to see if surrogates provide any benefit when including realistic cost data.

* genetic data
	+ AFLP data: next gen data would be better however the aim of a prioritization is to capture the main gradients of genetic variation--not all of the variation.
	+ adegenet: whilst other software packages are more popular (eg. Structure), we used adegenet because information theory is more reliable than an ad-hoc second derivative process for comparing models. Additionally, it is computationally infeasible to analyze run 27 species in Structure in a suitable period of time.

### Concluding remarks
* take-home messages
	+ surrogates are effective for single-species prioritizations in conservation planning
	+ however, surrogates don't provide any noticeable benefits when using a comprehensive set of species in a multi-species prioritization
	+ conservation planners are encouraged to use comprehensive sets of species

>>>>

## Introduction
<!--  main text -->
Biodiversity is under threat. In recent history, many species have gone extinct [@r270; @r272]. Today, thousands of species are listed as endangered or near threatened in the International Union for Conservation of Nature (IUCN) red list (\url{www.iucnredlist.org}). In the near future, species are expected to face even more threats [eg. climate change; @r236; @r243]. In response to this crisis, the overarching aim of conservation is to maximize the long-term persistence of biodiversity [@r255; @r67].

Protected areas are one of the hallmarks of modern conservation [@r440]. They buffer species from anthropogenic impacts and provide places for biodiversity processes to persist (eg. nutrient cycling and water purification). Additionally, protected areas set the stage for direct management actions to enhance species recovery (eg. habitat rehabilitation, translocation, and invasive species management). Since the resources available for conservation are limited, protected areas must be cost-effective. As a consequence, plans for protected areas (prioritizations) are often generated using decision support tools [eg. $\texttt{MARXAN}$; @r22; $\texttt{Zonation}$; @r429]. These tools often formulate the benefit that a given species receives from a solution based on the proportion of its range that is protected. However, the true benefit that a species might receive is dependent on many other additional factors.

@@

Recently, conservation scientists have begun investigating surrogates to generate more effective prioritizations without the overheads associated with genetic data. Since adaptive genetic variation is driven by selection pressures, environmental variables have been explored as potential surrogates for genetic variation [eg. @r247; @r2]. On the other hand, neutral variation arises from a reduction in gene flow, and so variables that predict reduction in gene flow between areas have been explored. For instance, the isolation by distance theory [IBD; @r259] predicts that populations further apart will have lower gene flow, and in turn have greater neutral variation among them. The use of distance-based surrogates has been found to improve prioritizations in an island network [@r200]. However the effectiveness of distance-based surrogates remains untested in contiguous systems conventionally used in conservation planning where gene flow is complicated by additional factors (eg. elevation in terrestrial systems). Before conservation planners can reliably use surrogates for genetic data in conservation planning, the effectiveness of these surrogates needs to be verified. Otherwise, if these surrogates are ineffective, conservation planners may waste precious resources on places with little benefit.

@@

## Methods
### Study system
To address the aims of the study, we utilized species distribution and genomic data collected by the IntraBioDiv project [@r451; see @r452 for methodological details]. This dataset was particularly well suited; it provides spatially explicit genomic data for a multitude of species across a broad geographic area over large environmental gradients. As part of this project, data were collected for 27 alpine plant species were across the European Alps according to a grid (22.3 km $\times$ 25 km; Figure S1). Briefly, in every second cell, plant samples were collected from three individuals for each species present inside the cell. Cells that did not contain any plant samples were omitted from our analysis. Samples were genotyped using amplified fragment length polymorphisms [AFLP; @r453]. Matrices denoting the presence/absence of polymorphisms at loci were constructed independently for each species. Thus for each second grid cell, for each of the species in the grid cell, the dataset contains information describing the genomic variation of individuals inside the grid cell. We used these grid cells as planning units to develop prioritizations to investigate the effectiveness of surrogates for the (AFLP) genetic variation. Most of our analyses described hereafter were performed using the \texttt{R} software package [version `r paste0(version$major, '.', version$minor)`; @r192].

### Landscape data
We aimed to investigate the effectiveness of climatic and geographic surrogates. To describe the location of each grid cell, we projected the grid into an equi-distant coordinate system ($\texttt{ESRI:102031}$), calculated their centroids, and extracted its two-dimensional coordinates. To describe the climatic variation among each cell, we obtained 19 bioclimatic variables [approx. 1 km resolution; @r33]. These layers and the grid cells were projected into an equal-area coordinate system ($\texttt{ESRI:102014}$). The bioclimatic variables were clipped to the extent of the grid and subject to a principle components analysis (PCA; Table S1). The first two principle components (PCs) cumulatively explained `r round(pca.DF[[4]][2],1)` % of the variation and were used to generate two new layers. We calculated the average value for each PC layer in each grid cell, and used these values to characterize the main themes of climatic variation among the grid cells (Figure S2). We used the total human population size inside each planning unit to represent its opportunity cost (obtained at 1 km resolution from the Global Rural-Urban Mapping Project; GRUMP V1; @rXXX).

### Outlier locus detection
To investigate the effectiveness of the surrogates for adaptive and neutral genetic variation, we first need to identify which of the sampled loci--if any--were under selection for each species. We used outlier locus detection methods [evaluated in @r454] to achieve this. The basic premise underpinning these analyses is that neutral loci are expected to exhibit a specific level of variation among populations ($F_{st}$), and loci that deviate from this expectation should be under selection. The advantage of these methods, in contrast with environmental association analyses [eg. r@XXX], is that they do not use environmental data to identify loci under selection, which would have biased our analysis. In an earlier study, Bothwell _et al._ [-@r455] applied such outlier detection methods to one of the species in this dataset (_Getiana nivalis_). Here, we adopted a similar methodology and applied it to each of the 27 species in the dataset.

@@

### Prioritizations
We used the unreliable representative and adequate prioritization (URAP) formulation of the reserve selection problem [@r465]. This formulation can identify the minimum number of planning units required to preserve a proportion of the species' range as well as a proportion of the species' variation using an attribute space [_sensu_ @r465]. We treated the genomic and surrogate variables as distinct attribute spaces. We set the demand points for each species and its attribute space using the values associated with the planning units. For every prioritization we generated, we calculated the proportion of the adaptive (if detected) and neutral genetic variation that every prioritization secured to assess its performance. We solved all reserve selection problems to within `r (gurobi.params.LST[[MODE]]$MIPGap*100)` % of optimality using the $\texttt{rapr}$ R package [@r465] and $\texttt{Gurobi}$ [@r463].

To understand how prioritizations generated using surrogate-based targets might perform in conservation planning scenarios, we generated three sets of prioritizations: a set of single-species prioritizations, a set of multi-species prioritizations, and a set of multi-species prioritizations that accommodated opportunity cost. For each of these sets, we generated prioritizations using only amount-based targets, prioritizations using both amount- and surrogate-based targets, and prioritizations using amount- and genetic-based targets. All amount-based targets were set as `r (rapr.params.LST[[MODE]]$amount.target*100)` % to ensure an adequate proportion of habitat was secured for each species [@rXXX]. All surrogate- and genetic-based targets were set as `r (rapr.params.LST[[MODE]]$surrogate.target*100)`. For each set, for each combination of targets, we identified the `r rapr.params.LST[[MODE]]$single.species.surrogate.replicates` solutions nearest to optimality for replication. Except for the single-species amount-only prioritizations for which we generated `r rapr.params.LST[[MODE]]$single.species.amount.replicates` replicates because they were associated with many optimal solutions with varying levels of effectiveness.

@@

## Statistical analysis
We tested for differences in the effectiveness of prioritizations representing typical scenarios using generalized linear mixed-effects models [GLMMs; using the $\texttt{lme4}$ R package; @r469]. We fit two full models to the proportion of genomic variation secured in the single- and multi-species prioritizations using a logit link. Both models contained a predictor variable indicating if the prioritization was generated using amount-only, surrogate, or genetic targets, a second predictor variable indicating if the genetic variation was adaptive or neutral, and the interaction between these variables. Additionally, both models contained an intercept-only random effect to ensure that comparisons were made within species. Each full model was subject to a step-wise term deletion routine using likelihood ratio tests (LRT) to assess term significance (terms were retained when $P < 0.05$). Finally, we conducted a post-hoc analysis using Tukey contrasts with a Bonferroni correction [using the $\texttt{multcomp}$ R package; @r470].

We investigated the relationship between surrogate-based targets and the genetic variation secured in a prioritization using nonlinear mixed-effects models (NLMMs; using the $\texttt{lme4}$ R package). Two models were fit to the adaptive and neutral genetic variation secured using a negative exponential formulation to estimate the rate of the accumulation of genetic variation ($y = a - (a - 0) \times exp(-r \times x)$; where $a$ is the asymptote and $r$ is the accumulate rate). Both models accommodated species-level random effects for the asymptote ($a$) and rate ($r$) parameters.

@@

<!-- Multi-species prioritizations -->
Environmental and geographic surrogates were not found to improve the performance of multi-species prioritizations (Figure 2; Figure 3). No significant differences in the effectiveness of prioritizations were found depending on whether they were generated using different targets (${\chi^{2}}_{`r multi.spp.AOV3[['Df']][1]`,`r multi.spp.AOV3[['Df']][2]`} = `r round(multi.spp.AOV3[['Chisq']][2],2)`$; $P `r pretty.pval(multi.spp.AOV3[['Pr(>Chisq)']][2])`$), if they were assessed using adaptive or neutral genetic variation (${\chi^{2}}_{`r multi.spp.AOV2[['Df']][1]`,`r multi.spp.AOV2[['Df']][2]`} = `r round(multi.spp.AOV2[['Chisq']][2],2)`$; $P `r pretty.pval(multi.spp.AOV2[['Pr(>Chisq)']][2])`$), nor the interaction between these variables (${\chi^{2}}_{`r multi.spp.AOV1[['Df']][1]`,`r multi.spp.AOV1[['Df']][2]`} = `r round(multi.spp.AOV1[['Chisq']][2],2)`$; $P  `r pretty.pval(multi.spp.AOV1[['Pr(>Chisq)']][2])`$).

<!-- Correlation analysis -->
The surrogate variables showed a greater utility for neutral genetic variation than adaptive variation (Figure 4). The accumulation rate ($r$) was `r round(summary(env.correlation.NLMM)$coefficients[1],2)` $\pm$ `r round(summary(env.correlation.NLMM)$coefficients[2],2)` S.E. for the proportion of adaptive variation captured in a prioritization using environmental surrogates. On the other hand, the rate ($r$) was `r round(summary(geo.correlation.NLMM)$coefficients[1],2)` $\pm$ `r round(summary(geo.correlation.NLMM)$coefficients[2],2)` S.E. for neutral variation using geographic surrogates. Thus minor increases in the target for either surrogate results in major increases in the benefit of the prioritization. However, the coefficient was much larger for neutral genetic variation so minor increases in the target can achieve substantial gains.

## Discussion
We aimed to investigate the effectiveness of environmental and geographic surrogates for genetic variation in conservation planning. This study is the first to empirically test this using a realistic system. We found that the inclusion of environmental and geographic surrogates in single-species prioritizations significantly improved the proportion of intra-specific genetic variation they secured. But we also found that the same surrogates did little to improve the multi-species prioritizations. Our results suggest that conservation planners can use a comprehensive set of species to secure a sufficient proportion of their intra-specific genetic variation.

### Single + multi-species analysis
* summarize results

* compare + contrast with other papers

* implication of results

### Correlation analysis
* summarize results

* compare + contrast with other papers

* implication of results

### Limitations/Future directions
We wish to alert practitioners to several limitations associated with our analysis. First, the size of the planning units we used (23 \times 25 km) is much larger than typically used in conservation planning scenarios (approx. 1--5 km^2). We used this resolution because the genetic data was collected at this scale. Whilst we could have interpolated the genetic data to a finer resolution and used smaller planning units, this would have biased our analysis because the surrogate variables are spatially auto-correlated. Second, we used geographic distances as surrogates for connectivity between planning units. Although connectivity between areas can be modeled using isolation by resistance measures derived from landscape data [eg. elevation; reviewed in @rXX], they are unlikely to be of use to conservation planners since they require genetic data to be tuned. Third, we used AFLP data to characterize the main gradients of genetic variation. Whilst next-generation sequencing provides higher resolution genetic information than AFLP data [reviewed in @rXXX], our methodology would still have utilised only the main gradients of the genetic variation to generate prioritizations in a feasible amount of time. Overall, the limitations associated with this study are relatively minor.

### Concluding remarks
The overarching aim of our study was to investigate the effectiveness of surrogates in conservation planning. We found that environmental and geographic surrogates are effective. However, they seem to be largely unnecessary in conservation planning scenarios because a comprehensive set of species can be used to achieve a similar result. Conservation planners are therefore urged to use a comprehensive set of species in prioritizations--one of the fundamental principles of conservation planning.
